Anxiety, avoidance, and sequential evaluation

### 2019-07-02 manuscripts/manuscript/manuscript.tex
co-author edits

\begin{abstract}
Anxiety disorders are characterized by a range of aberrations in the processing and response to threat, but there is little clarity what core pathogenesis might underlie these symptoms. Here we propose a decision theoretic analysis of maladaptive avoidance and embody it in a reinforcement learning model, which shows how a localized bias in beliefs can formally explain a range of phenomena related to anxiety. The core observation, implicit in standard decision theoretic accounts of sequential evaluation, is that avoidance should be protective: if danger can be avoided later, it poses no threat now. We show how a violation of this assumption --- a pessimistic, false belief that later avoidance will be unsuccessful --- leads to a characteristic propagation of fear and avoidance to situations far antecedent of threat. This single deviation can explain a surprising range of features of anxious behavior, including exaggerated threat appraisals, fear generalization, and persistent avoidance. Simulations of the model reproduce laboratory demonstrations of abnormal decision making in anxiety, including in situations of approach-avoid conflict and planning to avoid losses. The model also ties together a number of other seemingly disjoint issues in anxious disorders. For instance, learning under the pessimistic bias captures a hypothesis about the role of anxiety in the later development of depression. The bias itself offers a new formalization of classic insights from the psychiatric literature about the central role of maladaptive beliefs about control and self-efficacy in anxiety. This perspective is importantly different from previous computational accounts of beliefs about control in mood disorders, which neglected the sequential aspects of choice.
\end{abstract}

@@

Though anxiety disorders differ in their particular symptomology, and in the content and situations which elicit anxious responding, they all are similarly characterized by aberrations in the processing of and response to threat\citep{dsm5}. In particular, at least three symptoms are manifest across the anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted\citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, events, and thoughts\citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which often occurs well in advance of the materialization of actual threat\citep{Arnaudova2017}. Avoidance behaviors in anxiety are especially pernicious as they indirectly maintain anxiety by preventing learning from the non-occurrence of perceived threats. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

These symptoms are particularly puzzling from a decision theoretic perspective\citep{huys2015}. Fear and avoidance of situations far in the future violates the basic logic of evaluation over sequential trajectories of action. Put another way, distant threat need not necessarily impinge on decision making in the present. This is because avoidance is by nature protective: the ability to successfully avoid danger in the future means an agent need not also do so now. For instance, cars endanger pedestrians but can be reliably avoided by following traffic signals; given that, staying indoors offers little or no additional protection from accidents. This is an instance of a general property of evaluation in sequential decision making: the value of present action turns fundamentally on assumptions about subsequent events, which include the agent's own subsequent choices. Typically, it is appropriate to assume that an agent will continue make good (i.e. reward-maximizing/harm-minimizing) choices down the line.

This line of reasoning hints that a fundamental deficit in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of threat and excessive avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. Clinically and subclinically anxious individuals judge future threat as more likely than do non-anxious individuals\citep{ButlerMathews1983, ButlerMathews1987, MacleodByrne1996}. Importantly, the development and maintenance of clinical anxiety is strongly tied to diminished perceived control\citep{bandura1977, barlow2002, gallagher2014a}, such that anxious individuals are more likely to endorse the belief that they are unlikely or unable to mitigate future threat. Indeed, a belief (or lack thereof) of one's ability to successfully navigate future danger is associated with anxiety\citep{davey1996, dugas1997}, and an increased belief in perceived control over threat is correlated with symptom reduction across the family of anxiety disorders\citep{gallagher2014b}.

Here, we develop this idea into a model of evaluation under pessimistic assumptions about future choices. We show that a single, localized deviation can explain a surprising range of features of anxious behavior including exaggerated threat appraisal, fear generalization, and persistent avoidance. This account also offers a new formalization of classic insights from the psychiatric literature about the central role of beliefs about control and self-efficacy in anxiety \citep{bandura1977, barlow2002}. Specifically, we demonstrate a model with a misbelief in the reliability of future self-action gives rise to the anxious symptoms. Our perspective is importantly different from previous computational accounts of beliefs about control in mood disorders (e.g. \cite{HuysDayan2009}), which neglected the sequential aspects of choice.

@@

We model anxious decision making in the context of Markov decision processes (MDPs). A standard normative assumption is that agents attempt to optimize the expected cumulative discounted reward,

\begin{equation*}
Q^\pi(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \sum_{a'} \pi(a' \mid s') Q^\pi(s',a'),
\end{equation*}

but for any particular state-action $a$, this is necessarily defined only relative to a policy $\pi(a' \mid s')$ specifying the assumed distribution of \emph{future} choices. The return can be optimized self-consistently under the assumption that the agent makes the return-maximizing choice at each step in the future, leading to the familiar expression for the optimal values,

@@

The ``max'' operator in Eq.~\ref{eq:bellman} yields a fundamental asymmetry between approach and avoidance, as illustrated in Fig.~1. The assumption that the agent makes the return-maximizing choice at each step implies that opportunities for reward propagate recursively to earlier steps, but avoidable dangers do not. This principle is highlighted in a toy MDP: a deterministic open gridworld with terminal states containing a single reward and punishment (Fig.~1a). The optimal state values $V^* = \max_a Q^*(s,a)$ (Fig.~1b) reflect a ``mountain'' of opportunity propagating recursively from the reward. Conversely, because harm is avoidable in this environment, its negative value is contained: all states (even those adjacent to threat) represent the opportunity for reward.

An agent is not restricted to the return-maximizing assumption, and could anticipate encountering danger under different assumptions about the future. For example, an agent may expect to fail to take the correct protective actions (i.e., expects to use a nonoptimal $\pi(a \mid s)$), or may expect the world's future dynamics do not guarantee reliable avoidance even so (i.e., under stochastic or adversarial transition dynamics $P(s' \mid s,a)$). Consider an agent who has such pessimistic expectations about dangerous events at future steps. Computing returns with such pessimistic predictions might help to quantify uncertainty in outcomes (i.e. to learn different points in the distribution of possible returns)\cite{bellemare2017}, and choosing on this basis can be adaptive in that it increases the chances of safety in uncertain or even adversarial scenarios\cite{Garcia2015}.

Here we propose unrealistically pessimistic assumptions as a root cause of many anxious symptoms. Although such pessimism can be encoded either in the policy, $P(s' \mid s,a)$, or transition probabilities, $\pi(a \mid s)$, here for concreteness we work with distortions in the policy. In particular, we adopt the $\beta$-pessimistic value function from \cite{Gaskett2003}, to define an expectation of state-action value,

\begin{equation*}
Q^w(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \left( w \max_{a'} Q^w(s',a') + (1 - w) \min_{a'} Q^w(s',a') \right)
\end{equation*}

The parameter $w$ controls the degree of pessimism. An optimistic agent ($w = 1$) expects in the future to act fully in accordance with its preferences whereas a pessimistic agent ($w = 0$) expects to act contrary to its preferences. (We return to pessimism about state transitions in the discussion.) åFig.~1c,d illustrates the the consequences of different levels of pessimism for valuation in the example gridworld. An increasing expectation that the agent may fail to avoid in future causes the value of threat to permeate with increasing distance across the state space.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/01_field.png}}%
  }
  \par \textbf{Figure 1:} (A) A simple deterministic gridworld with two terminal states: one rewarding (blue) and one aversive (red). (B-D) States colored by their value under different levels of pessimism, with arrows showing an optimal trajectory. (B) For an optimistic agent ($w=1$), all states (other than the harmful state) take on positive value. (C) For a pessimistic agent ($w=0.5$), negative value spreads from the source to antecedent states. (D) With increasing pessimism ($w=0$), the extent of the spread grows worse. (Parameters: $\gamma = 0.95$)
\end{figure}

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/02_appavo.png}}%
  }
  \par \textbf{Figure 2:} (A) The Balloon Analog Risk Task (BART). The risk of balloon burst (point loss) increases with each pump. The optimal policy (number of pumps) under increasingly pessimistic valuation is presented for low (B) and high risk (C) balloons. The optimistic agent ($w=1$) prefers a policy reflecting the true environmental risk. The moderate ($w=0.6$) and strongly ($w=0.2$) pessimistic agents cash-out earlier, as is observed in anxious individuals. (D/E) By contrast, in the sleeping predator task the risk of loss is constant but the cost of loss increases. The value of reward pursuit under increasingly pessimistic valuation is presented for low (D) and high risk (E) scenarios. (Parameters: $\gamma = 1.0$)
\end{figure}

@@

The model can also explain findings of increased behavioral inhibition, measured as prolonged response times, in anxious individuals under threat\cite{bach2015}. In the behavioral inhibition task, participants seek tokens adjacent to a virtual "sleeping predator". Though the risk of predation is constant throughout a trial, the potential loss from capture increases with each token collected. Bach \cite{bach2015} finds that participants are slower to collect tokens as loss increases, and that this effect is modulated by subclinical anxiety. Our model recreates this phenomenon (Fig.~2d/e). Insofar that value and response time are negatively related \citep{need citation?}, and the relative value of approach as compared to avoidance is reduced under pessimistic assumptions about future action, then prolonged response times naturally follow.

@@

Our model predicts this result (Fig.~3b), though for a somewhat different reason. In our model, pessimistic (anxious) agents neglect large gains deeper in the tree, not because they fail to consider them, but because with increasing anxiety they increasingly expect the potential of choosing incorrectly afterward, thus failing to recoup the loss. This reinterpretation is consistent with the structure of the original task\citep{Huys2012, Lally2017}, where participants were required to plan in advance action sequences of two- to eight-steps (requiring the consideration of upwards of $2^8 = 256$ options). Uncertainty in one's ability to plan a sequence of reward-maximizing choices could similarly bias a participants towards smaller cost actions. Future research could use slight variants in the decision trees to tease apart these different interpretations.

% Include the larger, four-panel aversive pruning figure instead?

@@

So far, we have considered the asymptotic preferences implied the pessimistic value function (solved for through value iteration). But we can also consider learning under this value function (e.g., by Q-learning\citep{SuttonBarto2018} or DYNA\citep{sutton1991} using the $\beta$-pessimistic return), the dynamics of which may speak to the progression of symptoms.

@@

In addition to accounting for comorbidity of anxiety and depression, our model also hints at a reason for the longevity and recurrence of anxiety disorders even with treatment. Because pessimistic expectations allow for threat value to spread to states and actions far antecedent of the primary danger (e.g., Fig.~1d), it would accordingly also take a great many steps of iterative learning to correct all these exaggerated appraisals of threat. Frustratingly, these biased estimates of value may still remain even after a misbelief in the efficacy of future action is corrected for in a course of therapy. This may offer at least a partial answer to a classic puzzle in pathological avoidance, i.e. why it is so resistant to extinction \cite{moutoussis2018}, and to the unfortunately high rates of anxiety recurrence following treatment\citep{pittig2018}.

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/03_lh.png}}%
  }
  \par \textbf{Figure 4:} A simple deterministic gridworld with two terminal states: one rewarding (blue) and one aversive (red). (B-D) States colored by their value under different levels of pessimism, with arrows showing an optimal trajectory. For an optimistic agent ($w=1$), all states (other than the harmful state) take on positive value with learning. (E-G) For a pessimistic agent ($w=0.6$), negative value spreads from the source to antecedent states. As a result of avoidance, the agent learns reward is unobtainable and develops anergic symptoms (i.e. foregoes action). (Parameters: $\gamma = 0.95$)
\end{figure}

@@

In this final section, we offer a novel prediction regarding the free choice premium. Simply put, the free choice premium phenomenon predicts that, all else being equal, options which lead to more choice opportunities in the future are more valuable than those that lead to fewer future choice opportunities. A free choice premium has been observed in multiple behavioral experiments \cite{Leotti2010, ly2019}. A variant of a free choice premium paradigm from \cite{Leotti2011} and \cite{Leotti2014} is presented in Fig. 5a. In the task, participants repeated choose between a free choice option, allowing for an additional future choice, and a fixed choice option; importantly, both choices lead to identical, stochastic outcomes (e.g. 50/50 chance of [1, -1]). Previous studies have found a preference for the free choice option despite it conferring no additional benefits relative to the fixed choice option.

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/05_choice.png}}%
  }
  \par \textbf{Figure 4:} (A) The free choice premium paradigm from \cite{Leotti2011} and \cite{Leotti2014} with equal chance of outcome $R \in [1, -1]$. Non-anxious participants exhibit a preference for the free choice option (blue) despite it conferring no benefit over the fixed choice option (grey). (B) Pessimistic agents show an attenuated free choice bias. (Parameters: Q-learning with $\gamma = 1.0$, and inverse temperature, $\beta$, increased from 1 to 10 over 100 episodes. Each dot represents one of 500 simulated agents.)
\end{figure}

We (and others)\citep{ly2019} suspect the free choice premium reflects the usual assumption about choice (i.e. an agent will continue to make reward-maximizing choices in the future). Under such an optimistic assumptions, additional options are valuable insofar that they allow for additional future reward opportunities (or, at the worst, pose no possible harm). On the contrary, if anxiety reflects a violation of this typical assumption, then we should suspect anxious individuals will exhibit a diminished free choice bias. Indeed, the model makes this prediction (Fig. ~5b). We plan on explicitly testing this hypothesis in subclinical and clinical anxiety in the near future.

@@

Central to anxiety disorders are symptoms including exaggerated threat appraisal, threat generalization, and excessive avoidance. We have presented a simple computational account suggesting how a single underlying misbelief can give rise to these aberrations in learning and choice. Specifically, we showed how a failure to believe in the reliability of future self-action can effectively backpropagate negative value across states of the environment. This process results in a range of inferences and behaviors resembling those observed in clinical anxiety. Though this is by no means a complete account of anxiety, it ties together a surprisingly wide range of its symptoms.

One important ambiguity is that we have formalized pessimistic assumptions in terms of beliefs about the agent's own future actions: expecting failure in handling or avoiding future threat. Since Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$, pessimism can alternatively be encoded in this distribution (e.g., a belief that the world's response to one's choices is unpredictable or adversarial). Because the Bellman equation averages out both distributions, and because an unpredictable environment also reduces the efficacy of avoidance, either formulation can produce ultimately similar results. However from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs, which may be relevant, for instance, in guiding treatment.

Indeed, our account formalizes a longstanding strand of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, self-efficacy theory\citep{bandura1977} and the triple vulnerabilities model\citep{barlow2002} both posit that a reduced belief in the ability to effectively respond to future threat is involved in the genesis and maintenance of clinical anxiety. Similarly, the learned helplessness theories of anxiety\cite{alloy1990} claims clinical anxiety results from an uncertain belief in the controllability of the environment, such that future threat cannot be effectively mitigated or avoided. As we note above, the present model (though stated here in terms of self-efficacy) can accommodate either variant, and they are by no means mutually exclusive.

>>>>

\begin{abstract}
Anxiety disorders are characterized by a range of aberrations in the processing of and response to threat, but there is little clarity what core pathogenesis might underlie these symptoms. Here we propose a decision theoretic analysis of maladaptive avoidance and embody it in a reinforcement learning model, which shows how a localized bias in beliefs can formally explain a range of phenomena related to anxiety. The core observation, implicit in standard decision theoretic accounts of sequential evaluation, is that the potential for avoidance should be protective: if danger can be avoided later, it poses no threat now. We show how a violation of this assumption --- a pessimistic, false belief that later avoidance will be unsuccessful --- leads to a characteristic propagation of fear and avoidance to situations far antecedent of threat. This single deviation can explain a surprising range of features of anxious behavior, including exaggerated threat appraisals, fear generalization, and persistent avoidance. Simulations of the model reproduce laboratory demonstrations of abnormal decision making in anxiety, including in situations of approach-avoid conflict and planning to avoid losses. The model also ties together a number of other seemingly disjoint phenomena in anxious disorders. For instance, learning under the pessimistic bias captures a hypothesis about the role of anxiety in the later development of depression. The bias itself offers a new formalization of classic insights from the psychiatric literature about the central role of maladaptive beliefs about control and self-efficacy in anxiety. This perspective also extends previous computational accounts of beliefs about control in mood disorders, which neglected the sequential aspects of choice.
\end{abstract}

@@

Though anxiety disorders differ in their particular symptomology, and in the content and situations which elicit symptoms, they all are similarly characterized by aberrations in the processing of and response to threat\citep{dsm5}. In particular, at least three symptoms are manifest across the anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted\citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, events, and thoughts\citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which often occurs well in advance of the materialization of actual threat\citep{Arnaudova2017}. Such inapposite avoidance behaviors are especially deleterious aspects of the disoders, both because they interfere with daily life and because they indirectly maintain anxiety by preventing learning from the non-occurrence of perceived threats. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

These symptoms are particularly puzzling from a decision theoretic perspective\citep{huys2015}. In many circumstances, distant threat should not impinge upon decision making in the present. Indeed, fear and avoidance of situations far in the future violates the basic logic of evaluation over sequential trajectories of action. This is because avoidance is by nature protective: the ability to successfully avoid danger in the future means an agent need not also do so now. For instance, cars endanger pedestrians but can be reliably avoided by following traffic signals; given that, staying indoors offers little or no additional protection from accidents. This is an instance of a fundamental property of evaluation in sequential decision making: the value of present action turns fundamentally on assumptions about subsequent events, which include the agent's own subsequent choices. Typically, it is appropriate to assume that an agent will continue make good (i.e. reward-maximizing/harm-minimizing) choices down the line, and that choices at the current stage should reflect this.

This line of reasoning hints that a fundamental deficit in anxiety disorders may relate to this assumption, which otherwise should preclude the spread of threat to antecedent situations, and excessive avoidance. Indeed, anxiety disorders are associated with pessimistic beliefs about the future \citep{ClarkBeck2011}. Clinically and subclinically anxious individuals judge future threat as more likely than do non-anxious individuals\citep{ButlerMathews1983, ButlerMathews1987, MacleodByrne1996}. Importantly, the development and maintenance of clinical anxiety is strongly tied to diminished perceived control\citep{bandura1977, barlow2002, gallagher2014a}, such that anxious individuals are more likely to endorse the belief that they are unlikely or unable to mitigate future threat. Indeed, a belief (or lack thereof) of one's ability to successfully navigate future danger is associated with anxiety\citep{davey1996, dugas1997}, and an increased belief in perceived control over threat is correlated with symptom reduction across the family of anxiety disorders\citep{gallagher2014b}.

Here, we develop this idea into a formal model of evaluation under pessimistic assumptions about future choices. We show that a single, localized deviation can explain a surprising range of features of anxious behavior including exaggerated threat appraisal, fear generalization, and persistent avoidance. This account also offers a new formalization of classic insights from the psychiatric literature about the central role of beliefs about control and self-efficacy in anxiety \citep{bandura1977, barlow2002}. Specifically, we demonstrate a model with a misbelief about the reliability of future self-action gives rise to the anxious symptoms. In this way, our perspective also extends previous computational accounts of beliefs about control in mood disorders (e.g. \cite{HuysDayan2009}), which neglected the sequential aspects of choice.

@@

We model anxious decision making in the context of Markov decision processes (MDPs). A standard normative assumption is that agents attempt to optimize the expected cumulative discounted reward:

\begin{equation*}
Q^\pi(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \sum_{a'} \pi(a' \mid s') Q^\pi(s',a').
\end{equation*}

For any particular state-action $(s,a)$, this is necessarily defined only relative to a policy $\pi(a' \mid s')$ specifying the assumed distribution of \emph{future} choices. The return can be optimized self-consistently under the assumption that the agent makes the return-maximizing choice at each step in the future, leading to the familiar expression for the optimal values,

@@

The ``max'' operator in Eq.~\ref{eq:bellman} yields a fundamental asymmetry between approach and avoidance, as illustrated in Fig.~1. The assumption that the agent makes the return-maximizing choice at each step implies that opportunities for reward propagate recursively to earlier steps, but avoidable dangers do not. (To the extent obtaining reward or avoiding harm are only imperfectly attainable, this propagation and attenuation at each step are graded according to success probability, but the basic asymmetry remains.) This principle is highlighted in a toy MDP: a deterministic open gridworld with two terminal states, a rewarding state and a punishing one (Fig.~1a). The optimal state values $V^* = \max_a Q^*(s,a)$ (Fig.~1b) reflect a ``mountain'' of opportunity propagating recursively from the reward. Conversely, because harm is avoidable in this environment, its negative value is contained: all states (even those adjacent to threat) represent the opportunity for reward.

An agent is not restricted to the return-maximizing assumption, and could anticipate encountering danger under different assumptions about the future, including of course incorrect or pathological beliefs. For example, an agent may expect to fail to take the correct protective actions (i.e., expects to use a nonoptimal $\pi(a \mid s)$), or may expect the world's future dynamics do not guarantee reliable avoidance even so (i.e., under stochastic or adversarial transition dynamics $P(s' \mid s,a)$). Consider an agent who has such pessimistic expectations about dangerous events at future steps. Note that assumptions of this sort, even if incorrect, can serve adaptive purposes. It is common in machine learning theory, for instance, to optimize outcomes under worst-case assumptions. In general, this can help to ensure robustness and safety under uncertain or even adversarial scenarios \cite{Garcia2015}; related work in RL shows how computing returns with pessimistic predictions can help to quantify variability in outcomes (i.e. to learn different points in the distribution of possible returns)\cite{bellemare2017}, and explicitly take account of risk tolerance.

Here we propose unrealistically pessimistic assumptions as a root cause of many anxious symptoms. Although such pessimism can be encoded either in the policy, $P(s' \mid s,a)$, or transition probabilities, $\pi(a \mid s)$ (i.e. misbeliefs about either one's own avoidance actions or the environment's responses to them, a point we return to in discussion), here for concreteness we work with distortions in the policy. In particular, we adopt the $\beta$-pessimistic value function from \cite{Gaskett2003}, to define state-action value in expectation over a mixture of the best and worst action:

\begin{equation}\label{eq:pessimism}
Q^w(s,a) = r(s,a) + \gamma \sum_{s'} p(s' \mid s,a) \left( w \max_{a'} Q^w(s',a') + (1 - w) \min_{a'} Q^w(s',a') \right)
\end{equation}

The parameter $w$ controls the degree of pessimism. An optimistic agent ($w = 1$) expects in the future to act fully in accordance with its preferences, whereas a pessimistic agent ($w = 0$) expects to act contrary to its preferences. Fig.~1c,d illustrates the the consequences of different levels of pessimism for valuation in the example gridworld. An increasing expectation that the agent may fail to avoid in future causes the value of threat to propagate with increasing distance across the state space.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/01_field.png}}%
  }
  \par \textbf{Figure 1:} (A) A simple deterministic gridworld with two terminal states: one rewarding (blue) and one aversive (red). (B-D) States colored by their value under different levels of pessimism, with arrows showing an optimal trajectory. (B) For an optimistic agent ($w=1$), all states (other than the harmful state) take on positive value. (C) For a pessimistic agent ($w=0.5$), negative value spreads from the source to antecedent states. (D) With increasing pessimism ($w=0$), the extent of the spread grows worse, and the return-optimizing trajectory becomes more distorted and avoidant. (Parameters: $\gamma = 0.95$)
\end{figure}

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/02_appavo.png}}%
  }
  \par \textbf{Figure 2:} (A) The Balloon Analog Risk Task (BART). The risk of balloon burst (point loss) increases with each pump. The optimal policy (number of pumps) under increasingly pessimistic valuation is presented for low (B) and high risk (C) balloons. The optimistic agent ($w=1$) prefers a policy reflecting the true environmental risk. The moderate ($w=0.6$) and strongly ($w=0.2$) pessimistic agents cash-out earlier, as is observed in anxious individuals. (D/E) In the sleeping predator task, the risk of loss is constant but the cost of loss still increases as more rewards are gathered. The value of reward pursuit under increasingly pessimistic valuation is presented for scenarios with low (D) and high risk (E)  of predator awakening. The relative value of approach (vs. avoid) decreases with loss amount and threat level, and moreso under pessimistic assumptions. (Parameters: $\gamma = 1.0$)
\end{figure}

@@

The model can also capture findings of increased behavioral inhibition, measured as prolonged response times, in anxious individuals under threat\cite{bach2015}. In the behavioral inhibition task, participants seek tokens adjacent to a virtual "sleeping predator". Though in this variant the risk of predation is constant throughout a trial, the potential loss from capture still increases with each token collected. Bach \cite{bach2015} finds that participants are slower to collect tokens as loss increases, and that this effect is enhanced by subclinical anxiety. We can capture this effect in our model by noting that the relative value of approach compared to avoidance is reduced as the potential loss, and the risk the predator awakes, increases (Fig.~2d/e). These effects are amplified under more pessimistic (i.e., on our view, anxious) assumptions about future action. Thus, as before, anxiety in our model produces greater and earlier choice of avoidance (analogous to earlier cash-out in the BART). To extend this effect to reaction times, we adopt the further, standard assumption that actions (here, approach) are slower when their relative value compared to alternatives is lower \citep{oud2016}. In this case, the model captures behavioral inhibition (slower responses as threat increases), and its enhancement by anxiety, as measured by Bach. (Note that we need not assume the coupling of reaction times to action value spread is due to difficulty in decision formation per se, which Bach argues against: it may, for instance, reflect Pavlovian initiation biases \cite{niv2007}.)

@@

Our model predicts this result (Fig.~3b), though for a somewhat different reason. In our model, pessimistic (anxious) agents neglect large gains deeper in the tree, not because they fail to consider them (here we assume full evaluation of the Bellman equation), but because with increasing anxiety they increasingly expect the potential of choosing incorrectly afterward, thus failing to recoup the loss. This reinterpretation is consistent with the structure of the original task\citep{Huys2012, Lally2017}, where participants were required to plan in advance action sequences of two- to eight-steps (requiring the consideration of upwards of $2^8 = 256$ options). Uncertainty in one's ability to plan a sequence of reward-maximizing choices could similarly bias a participants towards smaller cost actions. Future research could use slight variants in the decision trees to tease apart these different interpretations.

% Include the larger, four-panel aversive pruning figure instead?
%%% I think this figure is OK but shouldnt it include optimal Q values for paths on the left also, showing why the best path is favored?
%%% also the graying out of the unchosen options on the right is maybe a little strong, hard to read, maybe darken these a bit (then you can also gray out 
%%% suboptimal paths on the left to make them really parallel?)

@@

So far, we have considered the asymptotic preferences implied the pessimistic value function (solved for directly by using value iteration to compute the return). But we can also consider learning under this value function (e.g., by variants of Q-learning\citep{SuttonBarto2018} or DYNA\citep{sutton1991} using the $\beta$-pessimistic return), the dynamics of which may speak to the progression of symptoms.

@@

In addition to accounting for comorbidity of anxiety and depression, our model also hints at a reason for the longevity and recurrence of anxiety disorders even with treatment. Because pessimistic expectations allow for threat value to spread to states and actions far antecedent of the primary danger (e.g., Fig.~1d), it would accordingly also take a great many steps of iterative learning to correct all these exaggerated appraisals of threat. Frustratingly, these biased estimates of value may still remain even after a misbelief in the efficacy of future action is corrected for in a course of therapy. This phenomenon (similar to failures of model-free RL algorithms to adjust to reward revaluation without extensive relearning; \cite{daw2005}) may offer at least a partial answer to a classic puzzle in pathological avoidance, i.e. why it is so resistant to extinction \cite{moutoussis2018}, and to the unfortunately high rates of anxiety recurrence following treatment\citep{pittig2018}.

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/03_lh.png}}%
  }
  \par \textbf{Figure 4:} A simple deterministic gridworld with two terminal states: one rewarding (blue) and one aversive (red). (B-D, E-G) The development of value expectancies over three steps of learning, for two levels of pessimism.
  States are colored by their value under different levels of pessimism, with arrows showing an optimal trajectory. (B-D) For an optimistic agent ($w=1$), all states (other than the harmful state) take on positive value with learning. (E-G) For a pessimistic agent ($w=0.6$), negative value spreads from the source to antecedent states. As a result of avoidance, the agent learns reward is unobtainable and develops anergic symptoms (i.e. foregoes action). (Parameters: $\gamma = 0.95$)
\end{figure}

@@

Finally, the model also offers a novel prediction tying anxious beliefs to a classic, but hitherto separate, phenomenon known as the free choice premium. This refers to the finding that, all else being equal, people tend to treat choice as itself valuable: i.e., choices which lead to more choice opportunities in the future are preferred to those that lead to fewer future choice opportunities. A free choice premium has been observed in multiple behavioral experiments \cite{Leotti2010, ly2019}. A variant of a free choice premium paradigm from \cite{Leotti2011} and \cite{Leotti2014} is presented in Fig. 5a. In the task, participants repeatedly choose between a free choice option, allowing for an additional future choice, and a fixed choice option; importantly, both choices lead to identical, stochastic outcomes (e.g. 50/50 chance of [1, -1]). Previous studies have found that human subjects (from a general, healthy population) prefer the free choice option despite it conferring no additional benefits relative to the fixed choice option.

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/05_choice.png}}%
  }
  \par \textbf{Figure 4:} (A) The free choice premium paradigm from \cite{Leotti2011} and \cite{Leotti2014} with equal chance of outcome $R \in [1, -1]$. Non-anxious participants exhibit a preference for the free choice option (blue) despite it conferring no benefit over the fixed choice option (grey). (B) Pessimistic agents show an attenuated free choice bias. The fractional preference for the free choice option over the simulated experiment is shown for three populations of subjects with levels of pessimism $w$ (y-axis; each dot represents an individual simulated agent, and the smoothed density of free choice bias for each pessimism level is also shown.)
  (Parameters: Q-learning with $\gamma = 1.0$, and inverse temperature, $\beta$, increased from 1 to 10 over 100 episodes.)
\end{figure}

On one account (e.g., \citep{ly2019}) this free choice preference directly and specifically reflects the assumption about sequential choice whose violation we argue is core to anxiety: i.e. that the agent will make reward-maximizing choices in the future. Under such an optimistic assumption (and given noisy and imperfect knowledge about the value of each option, due to learning), additional options are valuable in the sense that free choice can be expected to exploit the best among them: the maximum over several noisy values is, in general, larger than a single option from the same distribution. Our theory makes the novel prediction that, if as we hypothesize anxiety reflects a violation of this optimistic assumption, then anxious individuals will exhibit a diminished or reversed free choice bias, as shown in simulation in Fig. ~5b. Future research will be required to test this prediction.

@@

Central to anxiety disorders are symptoms including exaggerated threat appraisal, threat generalization, and excessive avoidance. We have presented a simple computational account suggesting how a single underlying misbelief can give rise to these aberrations in learning and choice. Specifically, we showed how a failure to believe in the reliability of one's future actions can effectively backpropagate negative value across states of the environment. This process results in a range of inferences and behaviors resembling those observed in clinical anxiety. Though this is by no means a complete account of anxiety, it ties together a surprisingly wide range of its symptoms.

One important ambiguity is that, for concreteness we have formalized pessimistic assumptions in terms of beliefs about the agent's own future actions: expecting failure in handling or avoiding future threat. Since Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$, pessimism can alternatively be encoded in this distribution (e.g., a belief that the world's response to one's choices is unpredictable or adversarial). Because the Bellman equation for the return averages over both distributions at each step, and because an unpredictable environment also reduces the efficacy of avoidance, either formulation can produce ultimately similar results in our simulations here. Other sorts of uncertainty (e.g. about the state of the environment, discussed below) can also produce similar effects for analogous reasons. However from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs, which may be relevant, for instance, in guiding treatment.

Indeed, our account formalizes a longstanding strand of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, self-efficacy theory\citep{bandura1977} and the triple vulnerabilities model\citep{barlow2002} both posit that a reduced belief in the ability to effectively respond to future threat is involved in the genesis and maintenance of clinical anxiety. Analogously, but focused less on the self, the learned helplessness theories of anxiety\cite{alloy1990} claims clinical anxiety results from an uncertain belief in the controllability of the environment, such that future threat cannot be effectively mitigated or avoided. As we note above, the present model and analysis (though stated here in terms of self-efficacy) can accommodate either variant. Also, they are by no means mutually exclusive, though they might also reflect multiple distinct (and differently treatable) routes to similar symptomatology.

Indeed, yet a third variant of our idea concerns the notion of state. In the present work, we considered only environments with fully observable states; in other words, there was no uncertainty regarding the state (e.g. grid world location) an agent currently occupied. In the real world, however, the state must be inferred from external and internal observations. This provides yet another source of uncertainty similar to the future policy and transition dynamics, which must be averaged over in evaluating actions \citep{kaelbling1998}. Thus another variant of our approach is to view anxiety as potentially also arising due to pessimistic biases in state inference\cite{Paulus2012}, which would again have similar effects on the resulting value function. For example, if an agent is erroneously biased toward believing herself more proximal to threat than is actually the case, then symptoms like exaggerated threat appraisal and avoidance should arise. Indeed, a bias in value functions could result in the diffuse spread of threat across objects\citep{norbury2018} and location\citep{schulz2018}, thereby causing otherwise benign states to be perceived as threatening.
